### 2020-04-03 Manuscript/manuscript.tex
small edits to address review comments

Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton.

@@

In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

@@

Accretionary orogenesis continued along the east margin of Laurentia with the arc-continent collision of the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprising amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}. This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}. The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.

@@

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton, as the record that lingers on the craton is similar. One interpretation is that there was successful break-up along the eastern margin during the ca. 580 to 550 Ma interval of rifting prior to the ca. 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent. The Maz-“Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting. Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca. 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca. 460 Ma Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.

@@

%The poles on this figure should be labeled
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Superior_Slave_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces. The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 25 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}}}
\label{fig:Superior_Slave_recons}
\end{figure}

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler of \cite{Maloof2006a} is used here. This Euler is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt Supergroup and the ca. 1.1 Ga Midcontinent Rift (Fig. \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similiarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2019a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2019a}.

Data from other terranes add resolution to the record. In particular, data from Greenland adds 12 poles between 1385 and 1160 Ma when there are only 4 poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire continent. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dykes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia \citep{Irving1972a}. Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dykes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dyke \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.

An exception to this overall agreement between poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Cenozoic Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts. There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite) however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

Another challenging portion of the Laurentia record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma. One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Bono2019a}. Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

Synthesizing the compilation of paleomagnetic poles for Laurentia into a composite path over the past 1.8 billion years presents a challenge given the highly variable temporal coverage. The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting spherical splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012a}. Applying such an approach can reduce the effect of spurious poles in regions of high data density where seeking to satisfy every mean pole position would result in jerky motion.

A synthesized pole path for Laurentia is developed here and used to develop a continuous paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles. The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}. This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example). One could use interpretations of paleogeographic connections with other cratons (e.g. Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself. In portions of the record with a more dense record of poles, such as ca. 1450, a calculated running mean is used to integrate constraints from multiple poles. This method follows the approach taken in the Phanerozoic (e.g. \citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps. When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model. Where there are no constraints a simple interpolation between constraints is made. While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain. These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca. 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}. The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China.

>>>>

Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton.

@@

The collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna in the paleogeographic model framework of \cite{Pehrsson2015a}. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

@@

Accretionary orogenesis continued along the east margin of Laurentia with the arc-continent collision of the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprising amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}. This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}. The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca. 1090 to 1030 Ma; \citealp{Rivers2008a}). Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca. 980 Ma \citep{Hynes2010a}.

@@

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton, as the record that lingers on the craton is similar. One interpretation is that there was successful break-up along the eastern margin during the ca. 580 to 550 Ma interval of rifting prior to the ca. 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent. The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting. Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca. 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca. 460 Ma Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.

@@

%The poles on this figure should be labeled
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Superior_Slave_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces. The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of narrower Manikewan with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 25 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}}}
\label{fig:Superior_Slave_recons}
\end{figure}

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} is used here. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt Supergroup and the ca. 1.1 Ga Midcontinent Rift (Fig. \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similiarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2019a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths such as ascending up to the Logan Loop, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2019a}.

Data from other terranes add resolution to the record. In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire continent. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dykes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia \citep{Irving1972a}. Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dykes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dyke \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1143 Ma lamprophyre dykes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.

An exception to this overall agreement between poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Cenozoic Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts. There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma. One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

Synthesizing the compilation of paleomagnetic poles for Laurentia into a composite path over the past 1.8 billion years presents a challenge given the highly variable temporal coverage. The method typically applied in the Phanerozoic is to develop synthesized pole paths either through fitting spherical splines through the data or calculating binned running means where the Fisher mean of poles within a given interval are calculated \citep{Torsvik2012a}. Applying such an approach can reduce the influence of spurious poles. Such synthesis is particularly important in regions of high data density where seeking to satisfy every mean pole position would result in jerky motion.

A synthesized pole path for Laurentia is developed here and used to develop a continuous paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles. The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}. This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example). One could use interpretations of paleogeographic connections with other cratons (e.g. Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself. In portions of the record with a more dense record of poles, such as ca. 1450 Ma, a calculated running mean is used to integrate constraints from multiple poles. This method follows the approach taken in the Phanerozoic (e.g. \citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps. When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model. Where there are no constraints a simple interpolation between constraints is made. While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain. These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca. 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}. The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China.
